---
title: "Credit Risk Model — Model Risk & Validation Report"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
params:
  confidence_level: 0.99
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(dplyr)
library(ggplot2)
library(knitr)

for (f in list.files(here::here("R"), pattern = "\\.R$", full.names = TRUE)) source(f)
```

## Executive Summary

This report documents the development, performance, and risk characteristics of the credit default prediction and portfolio loss simulation models. 
Two models were trained: a fully transparent logistic/WOE scorecard benchmark and an XGBoost model. Losses are simulated using a single-factor 
Gaussian copula Monte Carlo engine that accounts for correlated defaults across the portfolio, rather than a naive independent-default assumption.

## Data & Sample Definition

- Source: Lending Club accepted-loan data, restricted to matured loans (`Fully Paid` / `Charged Off`) so outcome labels are uncensored.
- Split: out-of-time (vintage-based), not random, train on earlier origination years, validate on the next year, and test on the most recent matured vintage, held out and touched once.

```{r data-summary}
train_df <- tar_read(train_df)
test_df  <- tar_read(test_df)

data.frame(
  Split = c("Train", "Test (OOT)"),
  `N loans` = c(nrow(train_df), nrow(test_df)),
  `Default rate` = c(mean(train_df$default_flag), mean(test_df$default_flag))
) %>% kable(digits = 4)
```

## Model Performance

```{r model-perf}
scorecard_eval <- tar_read(scorecard_eval)
xgb_eval <- tar_read(xgb_eval)

data.frame(
  Model = c("Logistic Scorecard", "XGBoost"),
  `Train AUC` = c(scorecard_eval$train_auc, xgb_eval$train_auc),
  `Test AUC` = c(scorecard_eval$test_auc, xgb_eval$test_auc),
  `Test KS` = c(scorecard_eval$test_ks, xgb_eval$test_ks),
  `PSI (train vs. OOT test)` = c(scorecard_eval$psi, xgb_eval$psi)
) %>% kable(digits = 3)
```

A PSI below 0.10 indicates no significant population shift between the
training and out-of-time test vintages; between 0.10-0.25 suggests moderate
drift worth investigating; above 0.25 indicates the model may need review or
refitting.

### Calibration (XGBoost, OOT test set)

```{r calibration-plot, fig.width=7, fig.height=4}
ggplot(xgb_eval$calibration, aes(x = mean_predicted_pd, y = observed_default_rate)) +
  geom_point(size = 2) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  labs(x = "Mean predicted PD (decile)", y = "Observed default rate",
       title = "Calibration: predicted vs. observed default rate") +
  theme_minimal()
```

## Portfolio Loss Simulation (Monte Carlo)

Correlated defaults are simulated using a single-factor Gaussian copula (Vasicek/ASRF-style), the same structure underlying the Basel IRB capital
formula, so the loss distribution reflects realistic tail risk rather than understating it as independent simulation would.

```{r mc-summary}
loss_distribution <- tar_read(loss_distribution)
risk_metrics_summary <- tar_read(risk_metrics_summary)

kable(risk_metrics_summary, digits = 0,
      col.names = c("Confidence", "Expected Loss", "VaR", "CVaR (Expected Shortfall)"))
```

```{r loss-hist, fig.width=7, fig.height=4}
var_line <- var_from_losses(loss_distribution, params$confidence_level)

ggplot(data.frame(loss = loss_distribution), aes(x = loss)) +
  geom_histogram(bins = 60, fill = "#2C3E50") +
  geom_vline(xintercept = var_line, color = "#E74C3C", linetype = "dashed", linewidth = 1) +
  annotate("text", x = var_line, y = Inf, label = paste0("VaR @ ", params$confidence_level * 100, "%"),
            vjust = 2, hjust = -0.1, color = "#E74C3C") +
  labs(x = "Simulated portfolio loss ($)", y = "Frequency",
       title = "Baseline simulated loss distribution") +
  theme_minimal()
```

## Stress Scenario

```{r stress}
stress_scenario <- tar_read(stress_scenario)

data.frame(
  Scenario = c("Baseline", "Stressed (systemic shock)"),
  `Expected Loss` = c(mean(stress_scenario$baseline$losses), mean(stress_scenario$stressed$losses)),
  `VaR (99%)` = c(
    var_from_losses(stress_scenario$baseline$losses, 0.99),
    var_from_losses(stress_scenario$stressed$losses, 0.99)
  ),
  `CVaR (99%)` = c(
    cvar_from_losses(stress_scenario$baseline$losses, 0.99),
    cvar_from_losses(stress_scenario$stressed$losses, 0.99)
  )
) %>% kable(digits = 0)
```

## Explainability

### Global feature importance (SHAP, XGBoost)

```{r shap-global, fig.width=7, fig.height=5}
shap_global <- tar_read(shap_global)

ggplot(head(shap_global, 15), aes(x = reorder(feature, mean_abs_shap), y = mean_abs_shap)) +
  geom_col(fill = "#2980B9") +
  coord_flip() +
  labs(x = NULL, y = "Mean |SHAP value|", title = "Top drivers of predicted default risk") +
  theme_minimal()
```

### Individual borrower example

```{r shap-local}
shap_values <- tar_read(shap_values)
explain_borrower(shap_values, row_index = 1, top_n = 6) %>%
  kable(digits = 4)
```

## 7. Limitations & Future Work

- LGD is approximated from historical recovery rates on charged-off loans and applied uniformly forward; a dedicated LGD model would improve precision.
- Asset correlation (`rho`) is a single portfolio-wide assumption; segment-specific correlations (by grade, purpose) could refine tail-risk estimates.
- This model targets post-funding risk (PD/LGD/EAD on already-approved loans). An acceptance/underwriting model with reject inference using the rejected-applications data is a natural extension, but out of scope here since rejected applications carry no outcome label.
